Exact Phylodynamics via Structured Markov Genealogy Processes

### 2023-07-13 ms.Rnw
small improvement to thm 2

Our first result will be an expression for the likelihood of a given pruned genealogy $P^*$ given the history $\Hist_t$ of the population process up to time $t=T(P^*)$.
Of course, not every pruned genealogy is compatible with $\Hist_t$.
In particular, $P^*$ is only compatible with $\Hist_t$ if every event in $P^*$ coincides with an event in $\Hist_t$.
That is, only if $\event(P^*)\subseteq{\event(\Hist_t)}$.
Moreover, certain events in $P^*$ may be incompatible with all possible histories.
For example, if $P^*$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*$ is incompatible with the population process itself.
Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(\eta,\eta')=1$ if and only if a change $\eta\to\eta'$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
More formally, $Q_u(\eta,\eta')=1$ if and only if there are $\omega\in\Omega$ and $t'<t\in\Rp$ such that $\U_{t'}(\omega)=u$, $\leftlim{Y}^{\prune(\Gen_t(\omega))}(t')=\eta$, and $Y^{\prune(\Gen_t(\omega))}(t')=\eta'$.
Using this, we define
\begin{equation*}
  \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
\end{equation*}
for $\xi\in\Xspace$, $\eta,\eta'\in\Yspace$, and $u\in\Jumps$.
Recall that $n:\Xspace\to\Zp$ is the deme-occupancy function, $r^u$ is the production of a jump of mark $u$, $\ell$ is the lineage-count function, and $s$ is the saturation.

\begin{thm}\label{thm:pruned_lik}
  Let $P^*$ be a given pruned genealogy, $t=T(P^*)$, and $y^*=Y^{P^*}$.
  Then
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}\,\times\,\prod_{e\in\event(\Hist_t)}{\phi_{U_e}\Big(X_{e},\yt^*(e),y^*(e)\Big)}.
  \end{equation*}
\end{thm}
\begin{proof}
  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*}=0$.
  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $\U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, one ends up with $s_i$ black balls, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Since our genealogies are labelled, we are interested in the probability that the specific lineages are involved.
  We therefore divide by $\binom{\ell_i}{s_i}$.
  Since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
  \end{equation*}
  Returning to the possibility that $P^*$ is incompatible with $\Hist_t$, since $\Prob{P^*}=0$ if either any $Q_u=0$ or of any event of $P^*$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

@@

\begin{thm}\label{thm:obsc_lik}
  Let $V^*$ be a given obscured genealogy and $t=T(V^*)$.
  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$ for all $u\in\Jumps$, $t\in\Rp$, $\xi,\xi'\in\Xspace$, and $\eta\in\Yspace$.
  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ for all $t,\xi,\xi'$ whenever $Q_u(\eta,\eta')=1$.
  Let $\rdm{y}_t$ be the Markov jump process generated by the kernel $\pi$, with jump times fixed at $\event(\Hist_t)$.
  Then
  \begin{equation*}
    \Prob{V^*\vert\Hist_t}=\Expect{\prod_{e\in\event(\Hist_t)}{\frac{\phi_{U_e}(X_e,\leftlim{\rdm{y}}_t,\rdm{y}_t)}{\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}},
  \end{equation*}
  the expectation being taken over $\rdm{y}_t$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \Prob{V^*\vert\Hist_t}=\int{\Indicator{\obs(P)=V^*}\,\Prob{P\vert\Hist_t}\,\dd{P}},
  \end{equation}
  the integral being taken over all pruned genealogies.
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $V^*$ as sample paths from a Markov process on $\Yspace$ and
  treat the integral in \cref{eq:IS1} as an expectation over these paths.
  Conditional on $\Hist_t$, the probability kernel $\pi$ generates a Markov chain, $\rdm{\hat{y}}_k$, on $\Yspace$ such that
  \begin{equation*}
    \Prob{\rdm{\hat{y}}_k=\hat{y}_k\vert\rdm{\hat{y}}_{k-1}=\hat{y}_{k-1}}=\pi_{U_{t_k}}(t_k,\leftlim{X}_{t_k},X_{t_k},\hat{y}_{k-1},\hat{y}_k).
  \end{equation*}
  $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
  The conditions on $\pi$ guarantee that for every $P$ such that $\obs(P)=V^*$, then $Y^P$ is a sample path of this process and, conversely, that for every sample path $y$, there is a pruned genealogy $P=P(y)$ such that $\obs(P)=V^*$ and $Y^P(t)=y_t$.

>>>>

Our first result will be an expression for the likelihood of a given pruned genealogy $P^*$ given the history $\Hist_t$ of the population process up to time $t=T(P^*)$.
Of course, not every pruned genealogy is compatible with $\Hist_t$.
In particular, $P^*$ is only compatible with $\Hist_t$ if every event in $P^*$ coincides with an event in $\Hist_t$.
That is, only if $\event(P^*)\subseteq{\event(\Hist_t)}$.
Moreover, certain events in $P^*$ may be incompatible with all possible histories.
For example, if $P^*$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*$ is incompatible with the population process itself.
Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(\eta,\eta')=1$ if and only if a change $\eta\to\eta'$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
More formally, $Q_u(\eta,\eta')=1$ if and only if there are $\omega\in\Omega$ and $t'<t\in\Rp$ such that $\U_{t'}(\omega)=u$, $\leftlim{Y}^{\prune(\Gen_t(\omega))}(t')=\eta$, and $Y^{\prune(\Gen_t(\omega))}(t')=\eta'$.
Recall that $n:\Xspace\to\Zp$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.

\begin{thm}\label{thm:pruned_lik}
  Let $P^*$ be a given pruned genealogy, $t=T(P^*)$, and $y^*=Y^{P^*}$.
  Then
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}\,\times\,\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y(e))}{r^{U_e}}{s(\yt(e),y(e))}\,Q_{U_e}(\yt(e),y(e))}.
  \end{equation*}
\end{thm}
\begin{proof}
  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*}=0$.
  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $\U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, one ends up with $s_i$ black balls, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Since our genealogies are labelled, we are interested in the probability that the specific lineages are involved.
  We therefore divide by $\binom{\ell_i}{s_i}$.
  Since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
  \end{equation*}
  Returning to the possibility that $P^*$ is incompatible with $\Hist_t$, since $\Prob{P^*}=0$ if either any $Q_u=0$ or of any event of $P^*$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

\begin{equation*}
  \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
\end{equation*}

@@

\begin{thm}\label{thm:obsc_lik}
  Let $V^*$ be a given obscured genealogy and $t=T(V^*)$.
  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$ for all $u\in\Jumps$, $t\in\Rp$, $\xi,\xi'\in\Xspace$, and $\eta\in\Yspace$.
  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ if and only if $Q_u(\eta,\eta')=1$.
  Let $\rdm{y}_t$ be the Markov jump process generated by the kernel $\pi$, with jump times fixed at $\event(\Hist_t)$.
  Then
  \begin{equation*}
    \Prob{V^*\vert\Hist_t}=\Expect{\prod_{e\in\event(\Hist_t)}{\frac{\scriptsize\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_t,\rdm{y}_t)}}{\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}},
  \end{equation*}
  the expectation being taken over $\rdm{y}_t$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \Prob{V^*\vert\Hist_t}=\int{\Indicator{\obs(P)=V^*}\,\Prob{P\vert\Hist_t}\,\dd{P}},
  \end{equation}
  the integral being taken over all pruned genealogies.
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $V^*$ as sample paths from a Markov process on $\Yspace$ and
  treat the integral in \cref{eq:IS1} as an expectation over these paths.
  Conditional on $\Hist_t$, the probability kernel $\pi$ generates a Markov chain, $\rdm{\hat{y}}_k$, on $\Yspace$ such that
  \begin{equation*}
    \Prob{\rdm{\hat{y}}_k=\hat{y}_k\vert\rdm{\hat{y}}_{k-1}=\hat{y}_{k-1}}=\pi_{U_{t_k}}(t_k,\leftlim{X}_{t_k},X_{t_k},\hat{y}_{k-1},\hat{y}_k).
  \end{equation*}
  $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
  The conditions on $\pi$ guarantee that for every $P$ such that $\obs(P)=V^*$, then $Y^P$ is a sample path of this process and, conversely, that for every sample path $y$, there is a pruned genealogy $P=P(y)$ such that $\obs(P)=V^*$ and $Y^P(t)=y_t$.
  In particular, they guarantee that $\pi_u(\xi,\xi',\eta,\eta')\propto{Q_u(\eta,\eta')}$.
